---
title: "Assignment 3"
author: "John Chen"
date: "March 01, 2024"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    theme: cerulean
    highlight: tango
    css: styles.css
    fig_caption: true
    self_contained: true
link-citations: true
---

# Introduction

This report aims to answer the question: When and where do most wild fires occur? The [Fire Program Analysis fire-occurrence database](https://www.kaggle.com/datasets/rtatman/188-million-us-wildfires?rvi=1) 1.88 million recorded wild fires in the US from 1992 to 2015. The interest for this investigation was sparked by an sudden increase of wild fires around the world. I wish to examine the pattern in frequencies, more importantly, the irregularities in frequency of occurrences. It is expected that frequencies are impacted by the occurrences of other environmental disasters, like rains and hurricanes. This report is only meant to gain basic insight into the frequencies of fires throughout a long period of time. 
This data set contains the discovery date of the fire, the acres burned, location of the fire, and the cause of fire. 

```{r echo=FALSE, warning=FALSE, message=FALSE}
packages <- c("tidyverse", "DBI", "RSQLite", "lubridate", "hms", "kableExtra","gridExtra","vtable","knitr", "broom", "leaflet","sf")
for(p in packages){
 if(!require(p, character.only = TRUE)){
  install.packages(p) 
 }
 library(p, character.only = TRUE)
}
```

```{r echo=FALSE}
db <- dbConnect(RSQLite::SQLite(), "FPA_FOD_20170508.sqlite")
query <- "SELECT * FROM Fires"
result <- dbGetQuery(db, query)
dbDisconnect(db)
```
```{r echo=FALSE, results='hide'}
wildfire <- subset(result, select=-Shape)
wildfire
summary_data <- summary(wildfire)
kable(t(summary_data), format = "markdown")
```
# Method
Based on Kaggle source, this data is collected from records of federal, state, and local fire organizations. Redundant or duplicated data were removed, to the degree possible. The data was simplified by reformating the Julian dates into calendar time series and individual columns of year, month, and day.
I then kept all variables that were irrelevant to the question of interest: the State that the fire was in, year of the fire, the discovery date of the fire, the discovery time of the fire, the cause code of the fire, the cause of the fire, latitude, longitude, contained date, contained time, fire size, fire size class, county.
I will create a variable for season based on discovery date to also examine which season has the most wild fire.

To examine patterns in wild fire frequency, I will make histogram of year to see if there is an increasing or decreasing pattern. If there is an increasing pattern, it supports the claim that frequency of wild fire increases throughout time(both year and season will be examined). Spatial plots will be used to see the location of wild fire hot spots in the US during the time that would be concluded as the worst fire season. 
A chi-square test will be conducted on causes and season to see if there are any correlation. If there are, we should identify the combination of the two that relatively causes the most fires.
Since wild fires is a good fit for poisson distribution, I will use GLM to examine which values of variables generates the most amount of wild fire.

# Results
```{r}
julian_origin <- lubridate::ymd("0000-11-25") - lubridate::years(4713)
wildfire$DISCOVERY_DATE <- as.Date(wildfire$DISCOVERY_DATE, origin = julian_origin)
wildfire$DISCOVERY_TIME <- as_hms(strptime(wildfire$DISCOVERY_TIME, "%H%M"))

wildfire$CONT_DATE <- as.Date(wildfire$CONT_DATE, origin = julian_origin)
wildfire$CONT_TIME <- as_hms(strptime(wildfire$CONT_TIME, "%H%M"))
```
```{r results='hide'}
wildfire <- wildfire[c("STATE","FIRE_YEAR","DISCOVERY_DATE","DISCOVERY_DOY","DISCOVERY_TIME", "STAT_CAUSE_DESCR","CONT_DATE","CONT_DOY","CONT_TIME", "FIRE_SIZE", "FIRE_SIZE_CLASS", "LATITUDE", "LONGITUDE", "COUNTY")]
summary(wildfire)
```
```{r echo=FALSE, results='hide'}
wildfire %>%
  filter(is.na(CONT_DATE)) %>%
  ggplot() + geom_bar(aes(x=FIRE_SIZE_CLASS)) + labs(main = "Missing contained date by fire size class")
wildfire <- wildfire %>%
  mutate(
    CONT_DATE = as.Date(ifelse(is.na(CONT_DATE), DISCOVERY_DATE, CONT_DATE)),
    CONT_DOY = ifelse(is.na(CONT_DOY), DISCOVERY_DOY, CONT_DOY)
  )
summary(wildfire)
```
```{r}
wildfire <- wildfire %>%
  mutate(
    SEASON = case_when(
      month(DISCOVERY_DATE) %in% c(12,1,2) ~ "winter",
      month(DISCOVERY_DATE) %in% 3:5 ~ "spring",
      month(DISCOVERY_DATE) %in% 6:8 ~ "summer",
      month(DISCOVERY_DATE) %in% 9:11 ~ "fall"
    )
  )
wildfire$SEASON <- factor(wildfire$SEASON, levels = c("spring", "summer", "fall", "winter"))
wildfire$STAT_CAUSE_DESCR <- factor(wildfire$STAT_CAUSE_DESCR)
wildfire$FIRE_SIZE_CLASS <- factor(wildfire$FIRE_SIZE_CLASS, levels = c("A", "B", "C", "D", "E", "F", "G"))

```
```{r fig.cap="Figure 1"}
wildfire %>%
  ggplot() +
  geom_bar(aes(x=FIRE_YEAR),position="identity", alpha=0.4) +
  labs(x="YEAR", title = "Frequency of Fire")
wildfire %>%
  ggplot() +
  geom_bar(aes(x=FIRE_YEAR,fill=STAT_CAUSE_DESCR),position="identity", alpha=0.4) +
  labs(x="YEAR", title = "Frequency throughout the Years by Fire Cause", fill="Cause")
```
The frequencies of fire has not increased throughout the years. There does not seem to be a pattern for any distribution. 

## Which combination of season and causation has the highest frequency?
```{r fig.width=10}
wildfire %>%
  ggplot() +
  geom_bar(aes(x=FIRE_YEAR),position="identity", alpha=0.4) +
  labs(x="YEAR", title = "Frequency of Fire") + 
  facet_grid(SEASON ~ STAT_CAUSE_DESCR) +
  theme(text = element_text(size = 7)) +
  coord_fixed(ratio = 2/800) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Lightning causes frequencies of fire to peak in the summer. This makes sense since the occurrence of thunderstorms are most frequent during summer. As opposed to winter where there's almost no thunderstorms. This can be underlying an independence between season and causations.
I will perform a chi-square test to test the independence of the two variables.
```{r}
con <- table(wildfire$STAT_CAUSE_DESCR, wildfire$SEASON)
matrix <- apply(as.matrix.noquote(con),2,as.numeric)
chi <- chisq.test(con)
tidy_chi <- tidy(chi)
tidy_chi <- tidy_chi %>% select(-method)
kable(tidy_chi, format = "markdown", col.names = c("chi-squared statistic", "p-value", "degrees of freedom")) %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(2, bold = TRUE, color = "red") %>%
  add_header_above(c("Pearson's chi square test"=3))
```
Based on a p-value smaller than 2.2e-16 from chi-sqaure, we can reject the null that season is independent from cause. This supports the fact that there is a relation between season and causation.
```{r}
wildfire %>%
  ggplot(aes(x=STAT_CAUSE_DESCR, fill=SEASON)) +
  geom_bar(position="dodge") +
  labs(title = "Causation by season", x = "Causes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The frequency of fire is the highest during summer caused by lightning.

## Cluster Map for Large Fires (100+ acres)
```{r message=FALSE, warning=FALSE}
sampled_wildfire <- wildfire %>% 
  filter(!FIRE_SIZE_CLASS %in% c("A","B","C")) %>%
  sample_n(10000)
leaflet(sampled_wildfire) %>%
  addTiles() %>%
  addMarkers(clusterOptions = markerClusterOptions())
```
I sampled 10,000 observations from the dataset and plotted them on a clustered map. There is a larger amount of fires concentrated as we get closer to the equator. As you zoom in, you can see that the fires are most concentrated around Texas, Dallas. This concludes that Dallas Texas has the most amount of fire. We can examine more to see what is causing the fires around Dallas. 
```{r}
Dallas_coord <- c(32.779167, -96.808891)
buffer <- st_buffer(Dallas_coord, dist=2.75)
wildfire_around_texas <- wildfire[st_within(cbind(wildfire$lat, wildfire$lon), buffer),]
```

# Conclusion